Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-07-16 MS/ms.Rmd
Fixed some comments / suggestions from Megan. Still a few to deal with, but will see what others think. Added a new figure 1 to see if we can visualise things better

  pacman::p_load(flextable, tidyverse, equatags)  
```

@@

Meta-analysis is a well-developed research field with a long history in statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017; @Glass2015]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is becoming increasingly more prominent in the field of comparative physiology. Formal meta-analyses are part of a systematic review and involve a systematic literature-based search, extraction of data from studies meeting specific inclusion criteria, and the subsequent calculation of effects sizes from summary statistics provided by papers. A meta-analysis should follow and report upon this entire process carefully to ensure the review process and data collation, are as transparent and reproducible as possible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolution (PRISMA-EcoEvo) protocol, @ODea2021]. Aside from providing an up-to-date overview of a research field there are three major goals of meta-analysis: 1) to provide an overall mean estimate of a treatment effect or relationship, 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies and 3) attempt to identify gaps and publication biases [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013; @Borenstein2019]. A critically important feature of a meta-analysis is its ability to account for sampling error [@Koricheva2013]. The inverse of sampling error variance is used to weight studies, accordingly, in the analysis and make overall estimates more precise. While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on estimates because sampling variance cannot be formally accounted for in the analysis, even if the overall mean itself is unbiased [@Morrissey2016a].

Despite its widespread adoption and well established methodological procedures, meta-analysis is often criticised for mixing 'apples and oranges' -- in other words, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010; @Glass2015]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. Indeed, study specific differences in comparative physiology are widespread. For example, experimental designs can vary in the temperatures that they apply or in the dosages of chemical modifiers used in experimental treatments [e.g., @WuSeebacher2020]. To the uninitiated, this concern might not be unfounded given that heterogeneity in effects is often high in ecological and evolutionary meta-analyses [@Senior2016], telling us that the effect size varies a great deal from one population to another. However, to many, the very fact that highly heterogeneous effects exist is a core strength of meta-analysis [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. Indeed, exploring the reasons for why effects are likely to vary across studies "...is the spice of life" [p.g., 519; @Cooperetal2019].

The goal of our review is to briefly overview the process of meta-analysis, making a case for why heterogeneity in a meta-analysis is a good thing for answering fundamental research problems and guiding research directions. We introduce the idea of 'nuisance heterogeneity', showing comparative physiologists that it can be comfortably dealt with by re-formalising many existing effect sizes measures and/or by using multilevel meta-regression models. Indeed, a combination of these approaches might often be ideal. We first conduct a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic practices -- determining what types of effect sizes are commonly used, how often effect size estimates are impacted by 'nuisance heterogeneity' and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. We then formalise alternative effect sizes and their associated sampling variance to provide comparative physiologists opportunities to explicitly incorporate treatment differences into the effect size estimates to ease their interpretation. We then describe how more complex treatment differences, such as dosage differences (which may have non-linear effects) can be accommodated using multilevel meta-regression models. We hope to show how expanding the meta-analytic toolkit for comparative physiologists will provide new opportunities to address global scale problems that attempt to understand how organisms will cope with rapidly changing environments in the future.

@@

Meta-analysis is always concerned with effect heterogeneity; in other words, the factors that drive differences in the direction and magnitude an of effect within and across studies [@Nakagawa2017; @Gurevitch2018; @Cooperetal2019; @Borenstein2019]. The concept of heterogeneity is vitally important because it tells us how general a specific conclusion is likely to be, how much of the variance we see is the result of real biology (and not just sampling variance), and provides a way to understand what the key biological factors are leading to an effect [@Nakagawa2017; @Borenstein2019]. There are a plethora of phenomena driving effect heterogeneity, some of which are interesting and others that are less so.

Generally, contributors to effect size heterogeneity fall within a number of broad categories. First, sampling variance contributes to heterogeneity; the variability driven by the inherent uncertainty in sampling a given population. We cannot sample the entire population and nor do we necessarily want to. As a consequence, we need to rely on a sub-sample to estimate a population parameter (e.g., correlation between two traits). How many independent samples we collect impacts our precision in estimating this 'true' effect [@Borenstein2019]. In meta-analysis we formally include sampling variance to remove this source of heterogeneity directly because it is known -- hence why having sampling variances for an effect is so important [@Gurevitch2018; @Koricheva2013]. Second, there are sociological factors impacting heterogeneity. These include attitudes towards publishing non-significant results. Low powered studies that do not find an 'effect' or that find an effect in the opposite direction often lead to what is called publication bias [@Jennions2013; @Nakagawa2021b; @Rothstein2005]. Third, there is real biology driving differences in effects. This is particularly true in comparative physiology where different species with varying habitats, life-histories, mating systems and reproductive modes are synthesised. As comparative physiologists, this is the stuff that gets us up every morning! Much variation could result from shared evolutionary history (i.e., phylogeny) or because of the different thermal environments species (or populations) occupy. We can model these biological factors to test predictions from hypotheses that we believe are key players impacting mean effect size and direction. This improves our understanding of the biological world around us and helps inform on future research directions. Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it is important we know and understand this as it can determine how experiments are designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' [following from @Cooperetal2019] -- factors we know vary, but are rather unsurprising to us. These might include differences in temperature, pH, dosages, water potential etc. It might already be clear that these factors will have an impact on the magnitude of an effect, but they might distract from the core interest of a meta-analysis. In some cases, however, the latter factors might actually not be 'nuisance' variables but the main interest of a synthesis. Regardless, in many cases they can result in a more complicated interpretation of an effect measure then what is desired.

All the various sources of heterogeneity (except sampling variance) all contribute (to varying extents) to the 'apples and oranges' problem often touted as being an issues with some meta-analyses (or quantitative syntheses more generally). Indeed, some of this criticism revolves around combining studies that vary in their quality (e.g., observational vs experimental). Combining standardised effects, while ignoring these sources of heterogeneity can result in the meaning of a meta-analysis not being particularly useful to many. However, many argue that this is just what we are interested in synthesizing [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. In the words of Gene Glass, the father of meta-analysis: "Of course it [meta-analysis] mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. @Glass2015's intention was to simply highlight that whether comparing apples and oranges is a good idea simply boils down to the specific question of interest, the population one wishes to make conclusions about and how heterogeneous effects actually are. If we are interested in gaining insight into 'fruit' then synthesising apples and oranges is the only thing that makes sense. Instead, if we are interested in generalising only to apples then mixing fruit will not be ideal. In addition, if we want to understand why different 'fruit' vary in their effects, then we can only ask this question by including all fruit [@Borenstein2019].

@@

All title, abstract and full text screening was done by two people (DWAN and ML). Any conflicts or disagreements were resolved by discussion and a consensus on a paper was reached. Out of the 417 papers, 79 full texts were screened and we included a total of 62 for final data extraction. For full details on the search strings used, dates searched, the number of papers found, and our PRISMA flowchart (along with reasons for exclusion at the full text stage) we refer readers to the *Supplementary Materials*. To address our questions, we categorised effect sizes as falling into the following categories: Response Ratio (lnRR), Standardised Mean Differences (SMD; e.g., Hedges d, g or Cohen's d), Correlation or Fisher's Z -transformed Correlation, Regression Slopes (Slopes), Risk Ratios, or whether the paper analysed the raw data or summary statistics directly (i.e., means/proportions) (which we label as raw/mean/proportion). For all other effects we labelled as 'other'. To determine whether effect sizes were likely to vary based on some underlying continuous variable or were dependent on the context in which traits were measured, we assessed whether authors: 1) acknowledged this explicitly in the paper or 2) accounted for it in their models. If it was not clear, two of us (DWAN & PP) discussed the paper, and made a decision. For example, trait means, such as ectothermic metabolic rate (i.e., raw/mean/proportion-based effect size), will vary depending on the temperature being measured, the magnitude of differences between two treatments (e.g, contrast-based effect sizes - SMD or lnRR) might depend on the temperature or dosage used in treatments, and the strength of relationship between two variables (e.g., correlation between behaviour and metabolism) might depend on the temperature context under which metabolic rate or behaviour were measured. Given the high heterogeneity in effects used in comparative physiology, we expected this to be common to most effect sizes, however, it has not been formally quantified. We categorised the types of statistical models fit in each study as: 1) fixed-effect meta-analysis; 2) random effects meta-analysis (i.e., intercept only model, but has random study effects); 3) multi-level meta-analysis (i.e., intercept only model, but has multiple random effect levels such as study, species, observation-level random effects); 4) meta-regression models (i.e., a multilevel or random effect model with fixed effects - including random effects); 5) weighted regression models, including mixed effect models; and 6) linear regression models (including linear and generalised linear mixed effect models, and phylogenetic least squares models). Notably, only model types 1-5 are weighted models.

@@

Quantitative syntheses in comparative physiology commonly analyse raw summary statistics (`r with(effects, effects[effect == "Means","prop"])*100`\% of studies), followed closely by Fishers transformed correlation coefficients (Zr) (`r with(effects, effects[effect == "Zr","prop"])*100`\% of studies), standardised mean differences (SMD) and log response ratios (lnRR) (`r with(effects, effects[effect == "SMD","prop"])*100`\% and `r with(effects, effects[effect == "lnRR","prop"])*100`\%, respectively; Figure \@ref(fig:fig1)A). In many cases, studies make use of multiple effect size measures. Approximately `r with(hetero, hetero[cont_moderator_variable == "yes","percent"])*100`\% of effect sizes used within the study were deemed to be impacted by 'nuisance heterogeneity' in some form or another (Figure \@ref(fig:fig1)B), and of the studies where this was the case, only `r with(hetero_dealt, hetero_dealt[cont_moderator_variable_modelled == "yes","percent"])*100`\% of the studies clearly dealt with it (Figure \@ref(fig:fig1)B); usually through meta-regression modelling. Overall, `r with(weighted, weighted[meta_analysis_weighted == "weighted model","percent"])*100`\% of studies used a weighted meta-analytic model, however, of the studies self described as 'meta-analyses', `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","percent"])*100`\% (N = `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","n"])` of `r with(weighted_meta, sum(weighted_meta[,"n"]))`) did not actually weight their meta-analytic mean. Many studies used 'Linear Regression' (including linear and generalised linear mixed effects models) as the main type of model fit to the data (Figure \@ref(fig:fig1)C). Of the studies using weighted meta-analytic models `r with(het_reported, sum(het_reported[1:2,"percent"])*100)`\% (N = `r with(het_reported, sum(het_reported[1:2,"n"]))` meta-analyses) did not report any measure of heterogeneity.

```{r, fig1, fig.cap="**Summary of meta-analytic practices in comparative physiology.** (A) Different types of effect measures commonly used in meta-analyses and their releative frequency of use. (B) how many meta-analyses where 'nuisance heterogenity' have the potential to impact effect sizes, and the proportion of studies explicitly dealing with nuisance heterogenity. (C) Frequency of analytical approaches used to analyse effect measures. Note that only 'Weighted Regression', 'Random Effects Models', 'Multilevel Metaregression Models' and 'Multilevel Meta-analysis Models' are weighted models."}
p1 = dat %>% dplyr::group_by(effect_size_type) %>% janitor::tabyl(effect_size_type) %>%
  dplyr::summarise(SMD = dplyr::if_else(grepl("SMD", effect_size_type), n, 0),
                  lnRR = dplyr::if_else(grepl("Response Ratio", effect_size_type), n, 0),
                    Zr = dplyr::if_else(grepl("correlation", effect_size_type), n, 0),
                 Other = dplyr::if_else(grepl("other|Q10", effect_size_type), n, 0),
                 Means = dplyr::if_else(grepl("mean/proportion/raw", effect_size_type), n, 0),
                 lnCVR = dplyr::if_else(grepl("lnCVR", effect_size_type), n, 0),
                 Slope = dplyr::if_else(grepl("slope", effect_size_type), n, 0)) %>% 
  colSums(.) %>% data.frame(effect = names(.), n = .) %>% mutate(prop = n / sum(n)) %>% arrange(effect) %>% 
  ggplot() + tm + 
  geom_col(aes(x = effect, y = prop), width = 0.5, fill = "black", colour = text.colour, size = 0.5) +
  scale_y_continuous(expand=c(0,0), limits = c(0,0.35)) +
  labs(x = "Effect Size",
       y = "Frequency") +
  coord_flip()

@@

Meta-analysis requires comparative physiologists to make use of effect sizes that are suitable for answering their specific question. The key features of all effect sizes used in meta-analyses are that they are: 1) statistical parameters that have been placed on a common scale so that they can be compared across studies and 2) that they have some associated measure of precision (i.e., sampling variance)[@Koricheva2013; @Borenstein2009; @Schmid2021]. Effect sizes in comparative physiology are quite variable; they can be the mean difference between two groups (e.g., difference in mean between a control and treatment group - see @WuSeebacher2020), the slope of a regression (e.g., between body size and metabolic rate), a correlation between two variables, or even the raw mean (or variance) itself (e.g., comparing CTmax or CTmin across species - see @Sunday2014). In many cases, multiple types of effect sizes might be used depending on the question. For example, studies interested in the effect of temperature on some physiological trait that have measured the trait across studies experimentally manipulating temperature could analyse the mean 'effect' by creating contrast-based effect sizes or by analysing the mean in each temperature treatment. The choice as to which one to use will boil down to the types of data available from primary studies, the specific question of interest and whether studies are observational or experimental in nature. Considering our temperature example above, if we are interested in the magnitude of change between temperature treatments, we may want to use a contrast-based effect size, such as the log response ratio $\ln{RR}$. If we are instead interested in how the mean of some physiological trait changes with temperature we may instead model the mean itself. In the latter scenario, so long as all the means are measured in the same units (e.g., ; or can be converted to the same units) then the mean may provide the desired answer to a given question. We provide details on different effect measures and associated sample sizes in Table \@ref(tab:tablea).

@@

As is clear from Table \@ref(tab:tablea) none of the effect measures control for 'nuisance heterogeneity'. For example, the magnitude of response between two treatment groups for lnRR or SMD might depend on just how different the temperatures, dosages, pH or water potential applied in each treatment. Even modelling just the mean often needs to control for various sources of 'nuisance heterogeneity'. For example, standard metabolic rate (SMR) is temperature dependent. As such, one needs to standardise the temperature measurement to compare across studies [@White2006; @Uyeda2017] (if, of course, temperature is not of interest in of itself). While we could of course model this explicitly, this may not always be desirable. It can often be useful to standardise the effect to both ease its interpretation and simplify modelling. In Box 1 we provide readers with guidance on how comparative physiologists can apply corrections to the effect measure itself. We highlight two examples in the box focusing mainly on generality. Below, we show using a common, well understood effect measure of interest to comparative physiologists, $Q_{10}$, how we can formalise a traditional effect size, such as $lnRR$, to capture this 'nuisance variation'.

#### *Comparing Changes in Mean Physiological Rates, $Q_{10}$*
A very common occurrence in comparative physiology is to manipulate the temperature organisms experience and measure some physiological rate (e.g., metabolic rate). Using normal effect size measures (see Table \@ref(tab:tablea)) would result in the effect sizes varying in the types of temperatures applied within and across across studies contributing to 'comparability' issues. As such, it is often of interest to standardize effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020], which describes the multiplicative change in physiological rates across a 10°C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not have associated sampling variances derived for $Q_{10}$ making it challenging to make use of the full power of weighted meta-analytic models. Interestingly, $Q_{10}$ can be seen as a variant of *lnRR*, meaning that we can derive a $Q_{10}$ based effect size and sampling error using the well known mathematical properties of *lnRR*. This opens up the meta-analytic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

Prior to showing how the relevant $Q_{10}$ effect size can be calculated it can be helpful to understand its similarities to *lnRR*. The *lnRR* described by @Hedges1999 and extended by @Lajeunesse2015 and can be calculated as follows [but see also @Senior2020]:

>>>>

  pacman::p_load(flextable, tidyverse, equatags, png, magick)  
```

@@

Meta-analysis is a well-developed research field with a long history in statistics, medicine, psychology and ecology [@Gurevitch2018; @Nakagawa2017; @Glass2015]. It has emerged as the 'gold-standard' across various disciplines for quantitative synthesis and is becoming increasingly more prominent in the field of comparative physiology. Formal meta-analyses are part of a systematic review and involve a systematic literature-based search, extraction of data from studies meeting specific inclusion criteria, and the subsequent calculation of effects sizes from summary statistics provided by papers. A meta-analysis should follow and report upon this entire process carefully to ensure the review process and data collation, are as transparent and reproducible as possible [i.e., see the Preferred Reporting Items for Systematic Reviews and Meta-Analyses in Ecology and Evolution (PRISMA-EcoEvo) protocol, @ODea2021]. Aside from providing an up-to-date overview of a research field there are three major goals of meta-analysis: 1) to provide an overall mean estimate of a treatment effect or relationship, 2) to quantify effect size variance and understand key drivers explaining differences in effects across studies and 3) attempt to identify gaps and publication biases [@Nakagawa2017; @Gurevitch1999; @Gurevitch2018; @Cooper2009; @Koricheva2013; @Borenstein2019]. A critically important feature of a meta-analysis is its ability to account for sampling error [@Koricheva2013]. The inverse of sampling error variance is used to weight studies in the analysis and make overall estimates more precise. While many meta-analyses might employ unweighted analyses, these studies will not gain greater precision on estimates because sampling variance cannot be formally accounted for in the analysis, even if the overall mean itself is unbiased [@Morrissey2016a].

Despite its widespread adoption and well established methodological procedures, meta-analysis is often criticised for mixing 'apples and oranges' -- in other words, mixing effects from studies that lack comparability [@Gurevitch2018; @Arnqvist1995; @Gallo1979; @Carpenter2020; @Stewart2010; @Glass2015]. Lack of comparability could result from studies differing in experimental design, species, measurement variables and sampling units, which can impact the generalisability of meta-analytic results [@Arnqvist1995; @Stewart2010]. Indeed, study specific differences in comparative physiology are widespread. For example, experimental designs can vary in the temperatures that they apply or in the dosages of chemical modifiers used in experimental treatments [e.g., @WuSeebacher2020]. To the uninitiated, this concern might not be unfounded given that heterogeneity in effects is often high in ecological and evolutionary meta-analyses [@Senior2016], telling us that the effect size varies a great deal from one population to another. However, to many, having highly heterogeneous effects "...is the spice of life" [p.g., 519; @Cooperetal2019] because it provides opportunities to explore the reasons for why effects vary within and across studies [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015].

The goal of our review is to briefly overview the process of meta-analysis, making a case for why heterogeneity in a meta-analysis is a good thing for answering fundamental research problems and guiding research directions. We introduce the idea of 'nuisance heterogeneity', showing comparative physiologists that it can be comfortably dealt with by re-formalising many existing effect sizes measures and/or by using multilevel meta-regression models. Indeed, a combination of these approaches might often be ideal. We first conduct a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic practices -- determining what types of effect sizes are commonly used, how often effect size estimates are impacted by 'nuisance heterogeneity' and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. We then formalise alternative effect sizes and their associated sampling variance to provide comparative physiologists opportunities to explicitly incorporate treatment differences into the effect size estimates to ease their interpretation. Finally, we describe how more complex treatment differences, such as dosage differences (which may have non-linear effects) can be accommodated using multilevel meta-regression models. We hope to show how expanding the meta-analytic toolkit for comparative physiologists will provide new opportunities to address global scale problems that attempt to understand how organisms will cope with rapidly changing environments in the future.

@@

Meta-analysis is always concerned with effect heterogeneity; in other words, the factors that drive differences in the direction and magnitude an of effect within and across studies [@Nakagawa2017; @Gurevitch2018; @Cooperetal2019; @Borenstein2019]. The concept of heterogeneity is vitally important because it tells us how general a specific conclusion is likely to be, how much of the variance we see is the result of real biology (not just sampling variance), and provides a way to understand what the key biological factors are leading to an effect [@Nakagawa2017; @Borenstein2019]. There are a plethora of phenomena driving effect heterogeneity, some of which are interesting and others that are less so.

Generally, contributors to effect size heterogeneity fall within a number of broad categories. First, sampling variance contributes to heterogeneity; the variability driven by the inherent uncertainty in sampling a given population. We cannot sample the entire population and nor do we necessarily want to. As a consequence, we need to rely on a sub-sample to estimate a population parameter (e.g., correlation between two traits). How many independent samples we collect impacts our precision in estimating the 'true' effect [@Borenstein2019]. In meta-analysis we formally include sampling variance to remove this source of heterogeneity directly because it is known -- hence why having sampling variances for an effect is so important [@Gurevitch2018; @Koricheva2013]. Second, there are sociological factors impacting heterogeneity. These include attitudes towards publishing non-significant results. Low powered studies that do not find an 'effect' or that find an effect in the opposite direction often lead to what is called publication bias [@Jennions2013; @Nakagawa2021b; @Rothstein2005]. Third, there is real biology driving differences in effects. This is particularly true in comparative physiology where different species with varying habitats, life-histories, mating systems and reproductive modes are synthesised. As comparative physiologists, this is the stuff that gets us up every morning! Much variation could result from shared evolutionary history (i.e., phylogeny) or because of the different thermal environments species (or populations) occupy. We can model these biological factors to test predictions from hypotheses that we believe are key players impacting mean effect size and direction. This improves our understanding of the biological world around us and helps inform on future research directions. Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it is important we know and understand this as it can determine how experiments are designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' [following from @Cooperetal2019] -- factors we know vary, but are rather unsurprising to us (Figure \@ref(fig:fig1)). These might include differences in temperature, pH, dosages, water potential etc. It might already be clear that these factors will have an impact on the magnitude of an effect, but they might distract from the core interest of a meta-analysis. In some cases, however, the latter factors might actually not be 'nuisance' variables but the main interest of a synthesis. Regardless, in many cases they can result in a more complicated interpretation of an effect measure than what is desired.

```{r, fig1, fig.cap="*Sources of effect size heterogeneity in meta-analyses*. Heterogeneity can come in various forms from effect measures varying as a result of shared evolutionary history (i.e., phylogeny), different trait types measured, and even sources of 'nuisance heterogeneity'. Nuisance heterogeneity can be factors, such as temperature differences impacting effect size measures, or dosages. We can control for these types of heterogeniety by reformalising effect sizes and/or using multilevel metaregression as we describe in our review"}
image <- image_read(here::here("figs", "Fig1.png"))
image
```

All the various sources of heterogeneity (except sampling variance) contribute (to varying extents) to the 'apples and oranges' problem often touted as being an issues with meta-analyses (or quantitative syntheses more generally). Indeed, some of this criticism revolves around combining studies that vary in their quality (e.g., observational vs experimental). Combining standardised effects, while ignoring these sources of heterogeneity can result in the meaning of a meta-analysis not being particularly useful. However, many argue that this is just what we are interested in synthesizing [@Cooperetal2019; @Borenstein2019; @Gurevitch2018; @Glass2015]. In the words of Gene Glass, the father of meta-analysis: "Of course it [meta-analysis] mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. @Glass2015's intention was to simply highlight that whether comparing apples and oranges is a good idea simply boils down to the specific question of interest, the population one wishes to make conclusions about and how heterogeneous effects actually are. If we are interested in gaining insight into 'fruit' then synthesising apples and oranges is the only thing that makes sense. Instead, if we are interested in generalising only to apples then mixing fruit will not be ideal. In addition, if we want to understand why different 'fruit' vary in their effects, then we can only ask this question by including all fruit [@Borenstein2019].

@@

All title, abstract and full text screening was done by two people (DWAN and ML). Any conflicts or disagreements were resolved by discussion and a consensus on a paper was reached. Out of the 417 papers, 79 full texts were screened and we included a total of 62 for final data extraction. For full details on the search strings used, dates searched, the number of papers found, and our PRISMA flowchart (along with reasons for exclusion at the full text stage) we refer readers to the *Supplementary Materials*. To address our questions, we categorised effect sizes as falling into the following categories: Response Ratio (lnRR), Standardised Mean Differences (SMD; e.g., Hedges d, g or Cohen's d), Correlation or Fisher's Z -transformed Correlation, Regression Slopes (Slopes), Risk Ratios, or whether the paper analysed the raw data or summary statistics directly (i.e., means/proportions) (which we label as 'raw'). For all other effects we labelled as 'other'. To determine whether effect sizes were likely to vary based on some underlying continuous variable or were dependent on the context in which traits were measured, we assessed whether authors: 1) acknowledged this explicitly in the paper or 2) accounted for it in their models. If it was not clear, two of us (DWAN & PP) discussed the paper, and made a decision about whether the effect measure was likely to be impacted. For example, trait means, such as ectothermic metabolic rate (i.e., raw/mean/proportion-based effect size), will vary depending on the temperature being measured, the magnitude of differences between two treatments (e.g, contrast-based effect sizes - SMD or lnRR) might depend on the temperature or dosage used in treatments, and the strength of relationship between two variables (e.g., correlation between behaviour and metabolism) might depend on the temperature context under which metabolic rate or behaviour were measured. Given the high heterogeneity in effects used in comparative physiology, we expected 'nuisance heterogeneity' to be common to most effect sizes, however, it has not been formally quantified. We categorised the types of statistical models fit in each study as: 1) fixed-effect meta-analysis; 2) random effects meta-analysis (i.e., intercept only model, but has random study effects); 3) multi-level meta-analysis (i.e., intercept only model, but has multiple random effect levels such as study, species, observation-level random effects); 4) meta-regression models (i.e., a multilevel or random effect model with fixed effects - including random effects); 5) weighted regression models, including mixed effect models; and 6) linear regression models (including linear and generalised linear mixed effect models, and phylogenetic least squares models). Notably, only model types 1-5 are weighted models because they weight effect sizes based on the inverse of their sampling variance or sample size.

@@

Quantitative syntheses in comparative physiology commonly analyse raw summary statistics (`r with(effects, effects[effect == "Means","prop"])*100`\% of studies), followed closely by Fishers transformed correlation coefficients (Zr) (`r with(effects, effects[effect == "Zr","prop"])*100`\% of studies), standardised mean differences (SMD) and log response ratios (lnRR) (`r with(effects, effects[effect == "SMD","prop"])*100`\% and `r with(effects, effects[effect == "lnRR","prop"])*100`\%, respectively; Figure \@ref(fig:figsurv)A). In many cases, studies make use of multiple effect size measures. Approximately `r with(hetero, hetero[cont_moderator_variable == "yes","percent"])*100`\% of effect sizes used within the study we deemed to be impacted by 'nuisance heterogeneity' in some form or another (Figure \@ref(fig:figsurv)B), but of the studies where this was the case, only `r with(hetero_dealt, hetero_dealt[cont_moderator_variable_modelled == "yes","percent"])*100`\% of the studies clearly dealt with it (Figure \@ref(fig:figsurv)B); usually through meta-regression modelling. Overall, `r with(weighted, weighted[meta_analysis_weighted == "weighted model","percent"])*100`\% of studies used a weighted meta-analytic model, however, of the studies self described as 'meta-analyses', `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","percent"])*100`\% (N = `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","n"])` of `r with(weighted_meta, sum(weighted_meta[,"n"]))`) did not actually weight their meta-analytic mean. Many studies used 'Linear Regression' (including linear and generalised linear mixed effects models) as the main type of model fit to the data (Figure \@ref(fig:figsurv)C). Of the studies using weighted meta-analytic models `r with(het_reported, sum(het_reported[1:2,"percent"])*100)`\% (N = `r with(het_reported, sum(het_reported[1:2,"n"]))` meta-analyses) did not report any measure of heterogeneity, which is unfortaunte given how important these measures are to interpreting effects.

```{r, figsurv, fig.cap="**Summary of meta-analytic practices in comparative physiology.** (A) Different types of effect measures commonly used in meta-analyses and their relative frequency of use. (B) how many meta-analyses where 'nuisance heterogenity' have the potential to impact effect sizes, and the proportion of studies explicitly dealing with nuisance heterogenity. (C) Frequency of analytical approaches used to analyse effect measures. Note that only 'Weighted Regression', 'Random Effects Models', 'Multilevel Metaregression Models' and 'Multilevel Meta-analysis Models' are weighted models."}
p1 = dat %>% dplyr::group_by(effect_size_type) %>% janitor::tabyl(effect_size_type) %>%
  dplyr::summarise(SMD = dplyr::if_else(grepl("SMD", effect_size_type), n, 0),
                  lnRR = dplyr::if_else(grepl("Response Ratio", effect_size_type), n, 0),
                    Zr = dplyr::if_else(grepl("correlation", effect_size_type), n, 0),
                 Other = dplyr::if_else(grepl("other|Q10", effect_size_type), n, 0),
                 Means = dplyr::if_else(grepl("mean/proportion/raw", effect_size_type), n, 0),
                 lnCVR = dplyr::if_else(grepl("lnCVR", effect_size_type), n, 0),
                 Slope = dplyr::if_else(grepl("slope", effect_size_type), n, 0)) %>% 
  colSums(.) %>% data.frame(effect = names(.), n = .) %>% mutate(prop = n / sum(n)) %>% arrange(effect) %>% 
  ggplot() + tm + 
  geom_col(aes(x = effect, y = prop), width = 0.5, fill = "black", colour = text.colour, size = 0.5) +
  scale_y_continuous(expand=c(0,0), limits = c(0,0.35)) +
  labs(x = "Effect Size",
       y = "Frequency") +
  coord_flip()

@@

Meta-analysis requires comparative physiologists to make use of effect sizes that are suitable for answering their specific question. The key features of all effect sizes used in meta-analyses are that they are: 1) statistical parameters that have been placed on a common scale so that they can be compared across studies and 2) that they have some associated measure of precision (i.e., sampling variance)[@Koricheva2013; @Borenstein2009; @Schmid2021]. Effect sizes in comparative physiology are quite variable; they can be the mean difference between two groups [e.g., difference in mean between a control and treatment group - see @WuSeebacher2020], the slope of a regression (e.g., between body size and metabolic rate), a correlation between two variables, or even the raw mean (or variance) itself [e.g., comparing CTmax or CTmin across species - see @Sunday2014]. In many cases, different effect sizes might be used depending on the question. For example, studies interested in the effect of temperature on some physiological trait that have measured the trait across studies experimentally manipulating temperature could analyse the mean 'effect' by creating contrast-based effect sizes or by analysing the mean in each temperature treatment. The choice as to which one to use will boil down to the types of data available from primary studies, the specific question of interest and whether studies are observational or experimental in nature. Considering our temperature example above, if we are interested in the magnitude of change between temperature treatments, we may want to use a contrast-based effect size, such as the log response ratio $\ln{RR}$. If we are instead interested in how the mean of some physiological trait changes with temperature we may instead model the mean itself. In the latter scenario, so long as all the means are measured in the same units (e.g., ; or can be converted to the same units) then the mean may provide the desired answer to a given question. We provide details on different effect measures and associated sample sizes in Table \@ref(tab:tablea).

@@

As can be seen in Table \@ref(tab:tablea) none of the effect measures control for 'nuisance heterogeneity'. For example, the magnitude of response between two treatment groups for lnRR or SMD might depend on just how different the temperatures, dosages, pH or water potential applied in each treatment. Even modelling just the mean often needs to control for various sources of 'nuisance heterogeneity'. For example, standard metabolic rate (SMR) is temperature dependent. As such, one needs to standardise the temperature measurement to compare across studies [@White2006; @Uyeda2017] (if, of course, temperature is not of interest in of itself). While we could model this explicitly, this may not always be desirable. It can often be useful to standardise the effect to both ease its interpretation and simplify modelling. In Box 1 we provide readers with guidance on how comparative physiologists can apply corrections to the effect measure itself. We highlight two examples in the box focusing mainly on generality. Below, we show using a common, well understood effect measure of interest to comparative physiologists, $Q_{10}$, how we can formalise a traditional effect size, such as $lnRR$, to capture this 'nuisance variation'.

#### *Comparing Changes in Mean Physiological Rates, $Q_{10}$*
A common experiment in comparative physiology is to manipulate the temperature organisms experience and measure some physiological rate (e.g., metabolic rate). Using the effect size measures in Table \@ref(tab:tablea) would result in the effect sizes varying with the temperatures applied within and across across studies contributing to 'comparability' issues. For example, an effect size for a temperature manipulation of 10°C would be larger than an effect size for a temperature manipulation of 5°C (see Figure \@ref(fig:fig1)). As such, it is common to standardize effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020], which describes the multiplicative change in physiological rates across a 10°C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. Currently, however, we do not have associated sampling variances derived for $Q_{10}$ making it challenging to make use of the full power of weighted meta-analytic models. Interestingly, $Q_{10}$ can be seen as a variant of *lnRR*, meaning that we can derive a $Q_{10}$ based effect size and sampling error using the well known mathematical properties of *lnRR*. This opens up the meta-analytic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

Prior to showing how the relevant $Q_{10}$ effect size can be calculated it is useful to understand its similarities to *lnRR*. The *lnRR* described by @Hedges1999 and extended by @Lajeunesse2015 and can be calculated as follows [but see also @Senior2020]:
